Fractional Types

### 2019-11-21 fracGC/p.tex
minor cleanups; intro; and now ready to tacke space section

\paragraph*{Outline.} The paper solves the ancilla problem in
reversible and quantum computation using a novel concept:
\emph{fractional types}. In the next section, we introduce the problem
of ancilla, motivate its importance, and explain the limitations of
current solutions. In Sec.~\ref{sec:pi}, we review existing work by
James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
introduces a reversible programming language built using isomorphism
of types. Although it is conceivable that the concept of fractional
types would apply to general-purpose languages (after some
adaptation), its natural technical definition is using symmetries
present in categorical model of type isomorphisms. We present such a
definition in Sec.~\ref{sec:dyn}: that definition captures the main
ideas of fractional types and partially solves a more general and more
expressive instance of the ancilla problem. To make the memory
management explicit, Sec.~\ref{sec:space} provides an operational
semantics that exposes memory management details using a realization
of fractional types as first-class garbage collectors. In
Sec.~\ref{sec:dep} we refine the type system to keep track of ancillae
values. Technically this requires the introduction of pointed types on
top of which are defined both a monad and co-monad of singleton
types. We conclude with a discussion of related work and a summary of
our results.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Space Operational Semantics}
\label{sec:space}

cardinality of pi type:
\PIMEMcard{}
enumeration:
\PIMEMenum{}
combinator preserve size:
\PIMEMcardeq{}
semantics for Pi:
\PIMEMstate{}
\PIMEMstep{}
semantics for PiFracDyn:
\PIFMEMstate{}
\PIFMEMstep{}

@@

% General allocation and de-allocation. No need to keep track of
% lifetime. Just make sure location is back to initial value. Safe to
% de-allocate because indistinguishable from freshly created. No pointer
% equality and aliases though. Concurrent: one thread allocates; gc
% thread meets someone else

>>>>

\paragraph*{Outline.} The paper solves the ancilla problem in
reversible and quantum computation using a novel concept:
\emph{fractional types}. In the next section, we introduce the problem
of ancilla, motivate its importance, and explain the limitations of
current solutions. In Sec.~\ref{sec:pi}, we review existing work by
James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
introduces a reversible programming language built using isomorphism
of types. Although it is conceivable that the concept of fractional
types would apply to general-purpose languages (after some
adaptation), its natural technical definition exploits symmetries
present in categorical model of type isomorphisms. We present such a
definition in Sec.~\ref{sec:dyn} which captures and illustrates the
main novelties of fractional types and their runtime representations
as first-class gc processes. This presentation partially solves a more
general and more expressive instance of the ancilla problem as
illustrated with examples. To make the arguments regarding memory
management explicit, Sec.~\ref{sec:space} presents an abstract machine
with a heap component that grows and shrinks as ancillae are allocated
and de-allocated via fractional gc processes. In Sec.~\ref{sec:dep} we
lift ``plain'' programs manipulating ancillae to the type system in
order to exploit the power of dependent types to guarantee, for the
first time, the condition necessary for safe de-allocation of
ancillae. Technically this solution requires the introduction of
pointed types on top of which are defined both a monad and co-monad of
singleton types. The tracking of ancillae value is done within the
type system via the singleton types. A number of examples exploiting
the singleton-based type system are present to demonstrate the full
expressiveness of fractional types. We furthermore demonstrate that,
from the proofs of type safety, plain (but guaranteed to be safe)
programs can be extracted and executing without errors or runtime
checks in the conventional type system. The last section concludes
with a discussion of related work and a summary of our results.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Space-Tracking Abstract Machine}
\label{sec:space}

cardinality of pi type:
\PIMEMcard{}

enumeration:
\PIMEMenum{}

combinator preserve size:
\PIMEMcardeq{}

semantics for Pi:
\PIMEMstate{}
\PIMEMstep{}

semantics for PiFracDyn:
\PIFMEMstate{}
\PIFMEMstep{}

@@

General allocation and de-allocation. No need to keep track of
lifetime. Just make sure location is back to initial value. Safe to
de-allocate because indistinguishable from freshly created. No pointer
equality and aliases though. Concurrent: one thread allocates; gc
thread meets someone else
